Short-term neuronal and synaptic plasticity act in synergy for deviance detection in spiking networks

### 2023-01-12 paper-1/manuscript.tex
condition -> sequence for dev/msc/std

To show that our model is capable of deviance detection, we first show the responses of the sample network in all conditions (\dev{}, \msc{}, and \std{} sequences) in Figure \ref{fig2}a. Here, each row corresponds to one trial, aligned to stimulus input at t = 0, with the number of spikes in each millisecond bin indicated by color. All responses share an early activity pattern, which derives from the stimulated neurons and their immediate postsynaptic targets, but differences soon start to emerge. The \dev{} sequence responses, shown at the top of the plot, exhibit a clear peak in activity around 15 ms after stimulation. The \std{} sequence responses show a later, more diffuse activity peak around 20 ms, and the \msc{} sequence responses lack any obvious activity cluster. We sum these patterns up in Figure \ref{fig2}b, where we show the mean network activity, in terms of spikes per millisecond bin, across trials of each type. Clearly, the target response in the \dev{} sequence is greater than either the \msc{} or the \std{} response.

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the average spike count per neuron per target trial for each network and stimulus assignment, and compared these across conditions. Figure \ref{fig2}c shows a summary of the average \dev{} and \std{} response sizes, normalized by the response in the \msc{} sequence. The data clearly confirmed that \dev{} responses were larger than \msc{} (T = 6.38e+03, p = 2.75e-13; one-sided Wilcoxon signed-rank test, n = 120), and \std{} responses smaller (T = 687, p = 6.42e-15). Accordingly, the DDI (Figure \ref{fig2}d) was positive in most networks (T = 5.97e+03, p = 4.75e-10, median 0.114). This shows that our model was well behaved and suitable for an investigation of deviance detection.

@@

To understand why the same stimulus A evoked a higher response in \dev{} than in \msc{} with TA only, we first compared the average responses in the two conditions, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in Figure \ref{fig4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both conditions. The contrast (right) revealed a clear increase in \dev{} spikes, particularly in the later parts of the response, as shown by the predominance of orange areas, indicating higher \dev{} activity. Since the stimulus was identical in both conditions, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower $V_{TA}$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and conditions. Left and middle column, target trials in \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (\dev{} - \msc{}) in $V_{TA}$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Network-wide median contrast (\dev{} - \msc{}) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (two-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

@@

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two conditions. In order to assess the relative contribution of $V_{TA}$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $V_{TA}$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $V_{TA}$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $V_{TA}$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any condition. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between the non-target response contrast (\dev{} - \msc{}) and the contrast in $V_{TA}$ at the start of B trials, and associated linear regression across all excitatory neurons. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
\textbf{c} Median non-target response contrast (\dev{} - \msc{}) among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labeled "late", across networks and stimuli.
\textbf{c} Median contrast (\dev{} - \msc{}) in $V_{TA}$ among the early and late neurons, respectively, across networks and stimuli.
\textbf{d} Relationship across networks and stimuli between the median contrast in $V_{TA}$ among the early portion and the median contrast in the response to stimulation in B, and associated linear regression

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we found, for each neuron, the minimum response time (i.e., time to first spike) across B trials in \dev{}, \msc{} and \std{} sequences, and ranked neurons by this response time. The ranking in the sample network is shown in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all conditions separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast (\dev{} - \msc{}) to the contrast in $V_{TA}$ at the start of B trials. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in Figure \ref{fig5}c, where we correlated the same response contrast to $\Delta V_{TA}$ before A. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the \dev{} sequence, while the response of the later portion of the network decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

@@

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target \dev{} trials, relative to \msc{}, whereas differences between target \dev{} and \msc{} trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across \dev{} and \msc{} conditions as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both conditions. The \msc{} sequence, however, clearly evoked greater depression, as borne out in the contrast (\dev{} - \msc{}), which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both conditions. In the \msc{} sequence, we can attribute the difference between A and B primarily to different response patterns: The context, and thus $x$, was approximately the same, but responses to B tended not to spread out as much in this network. However, in the \dev{} sequence, the B response was limited almost exclusively to a very local patch around the stimulation site, with depression therefore similarly spatially limited. As a result, the contrast (\dev{} - \msc{}) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicating greatly increased synaptic depression in \dev{}.

>>>>

To show that our model is capable of deviance detection, we first show the responses of the sample network in all sequences (\dev{}, \msc{}, and \std{}) in Figure \ref{fig2}a. Here, each row corresponds to one trial, aligned to stimulus input at t = 0, with the number of spikes in each millisecond bin indicated by color. All responses share an early activity pattern, which derives from the stimulated neurons and their immediate postsynaptic targets, but differences soon start to emerge. The \dev{} sequence responses, shown at the top of the plot, exhibit a clear peak in activity around 15 ms after stimulation. The \std{} sequence responses show a later, more diffuse activity peak around 20 ms, and the \msc{} sequence responses lack any obvious activity cluster. We sum these patterns up in Figure \ref{fig2}b, where we show the mean network activity, in terms of spikes per millisecond bin, across trials of each type. Clearly, the target response in the \dev{} sequence is greater than either the \msc{} or the \std{} response.

We could readily identify similar patterns emerging across most networks and stimulus assignments. To show that deviance detection is robust to network and stimulus identity, we calculated the average spike count per neuron per target trial for each network and stimulus assignment, and compared these across sequences. Figure \ref{fig2}c shows a summary of the average \dev{} and \std{} response sizes, normalized by the response in the \msc{} sequence. The data clearly confirmed that \dev{} responses were larger than \msc{} (T = 6.38e+03, p = 2.75e-13; one-sided Wilcoxon signed-rank test, n = 120), and \std{} responses smaller (T = 687, p = 6.42e-15). Accordingly, the DDI (Figure \ref{fig2}d) was positive in most networks (T = 5.97e+03, p = 4.75e-10, median 0.114). This shows that our model was well behaved and suitable for an investigation of deviance detection.

@@

To understand why the same stimulus A evoked a higher response in \dev{} than in \msc{} with TA only, we first compared the average responses in the two sequences, sorting neurons by their response onset to permit a high-resolution view of the stimulus response as it propagates through the network. As shown in Figure \ref{fig4}a (left and middle), we found that the response developed in a qualitatively similar fashion in both sequences. The contrast (right) revealed a clear increase in \dev{} spikes, particularly in the later parts of the response, as shown by the predominance of orange areas, indicating higher \dev{} activity. Since the stimulus was identical in both sequences, and TA was the only plasticity mechanism in play, these differences had to be due to (1) different firing thresholds, and/or (2) different synaptic currents as a consequence of different prior activity within the trial.

\begin{figure*}%
    \centering
    \includegraphics{fig4.pdf}
    \caption{}
    \label{fig4}
\end{figure*}
\textbf{Fig. 4} Responses to \dev{} target trials are larger due to lower $V_{TA}$.
\textbf{a} Target trial average post-stimulus spike histogram in the sample network, showing trial time along the horizontal axis, and neurons along the vertical, sorted by the time of the first recorded spike across all trials and sequences. Left and middle column, target trials in \dev{} sequence and \msc{} sequence, respectively; right column, contrast between these two.
\textbf{b} Target trial average of the threshold adaptation voltage $V_{TA}$, with neuron order and columns as in panel a.
\textbf{c} Target trial average of the membrane potential $V_m$, with neuron order and columns as in panel a. The first 10 ms of the neurons under direct stimulation are masked out to avoid saturation of the colormap.
\textbf{d} Relationship across excitatory neurons of the sample network between the contrast (\dev{} - \msc{}) in $V_{TA}$ at the start of target trials and the contrast in average number of spikes fired in target trials, and associated linear regression.
\textbf{e} Network-wide median contrast (\dev{} - \msc{}) in $V_{TA}$ across networks and stimuli, plotted against the Pearson correlation coefficients $\rho$ of the relationship shown in the previous panel ($\Delta V_{TA}$ vs response size contrast across excitatory neurons). Each marker corresponds to one network and target stimulus. Red markers indicate a significant negative correlation (two-sided Wald test, significance level 0.05). Inset histograms show the distribution of the corresponding values, with orange portions indicating the 95\% confidence interval of the median, calculated by bootstrap with 10000 iterations.
\textbf{f} In the sample network, contribution of the contrast in $V_{TA}$ and $V_m$ to bins with increased firing in \dev{} (i.e., positive/red bins in the spike probability contrast, panel a), weighted by that same contrast.
\textbf{g} Contribution of $V_{TA}$ relative to the sum of the absolute values of the weighted contributions in panel f.
\textbf{h} Relative contribution of $V_{TA}$ across networks and stimuli, showing median (solid line) and inter-quartile range (shaded area)

@@

Yet, as we saw in Figure \ref{fig4}c, the membrane voltages also differed between the two sequences. In order to assess the relative contribution of $V_{TA}$ to the increased \dev{} response, we chose to focus exclusively on time points with higher firing probability in the \dev{} sequence, discarding data points with equal or lower firing probability relative to the \msc{} sequence. We then weighted the remaining contrasts in $V_{TA}$ and $V_m$ by the magnitude of the response difference, and summed across neurons. The resulting time courses, shown in Figure \ref{fig4}f, show how much these two measures contributed to the increased \dev{} response. While thresholds were slightly lowered throughout the response, the membrane potential was dramatically increased. In relative terms (Figure \ref{fig4}g), $V_{TA}$ constituted a major driver of increased firing right at the onset of the response, as well as briefly towards the end of the response. A similar pattern emerged across networks and stimuli, as shown in Figure \ref{fig4}h. Here, only the initial response was driven mainly by $V_{TA}$, while later differences were largely explained by differences in $V_m$, leading to a relative TA contribution of less than 20\% on average. This indicates that adaptation initiated and dominated the increase in firing at the start of trials, whereas the late response was enlarged primarily as a result of increased presynaptic excitation.

@@

\begin{figure*}%
    \centering
    \includegraphics{fig6.pdf}
    \caption{}
    \label{fig6}
\end{figure*}
\textbf{Fig. 6} Dev non-target responses are reduced due to adaptation in early responders to B.
\textbf{a} The sample network, colored by latency rank, i.e. the rank order of the time to first spike in response to stimulation at site B in any sequence. The response develops asymmetrically due to structural properties, not due to interactions with A. Note that non-responding neurons are ranked last and appear dark blue.
\textbf{b} Relationship across excitatory neurons of the sample network between the non-target response contrast (\dev{} - \msc{}) and the contrast in $V_{TA}$ at the start of B trials, and associated linear regression across all excitatory neurons. Neurons are colored by latency rank as in panel a. Additionally, the earliest 100 responders are shown in the left plot, while later responders are shown in the right plot. Inset histograms show the distributions of the plotted values, with the confidence intervals of their medians, calculated by bootstrap with 10000 iterations, in darker color.
\textbf{c} Median non-target response contrast (\dev{} - \msc{}) among the first 100 neurons to respond to stimulation in B, labeled "early", and among the later neurons, labeled "late", across networks and stimuli.
\textbf{c} Median contrast (\dev{} - \msc{}) in $V_{TA}$ among the early and late neurons, respectively, across networks and stimuli.
\textbf{d} Relationship across networks and stimuli between the median contrast in $V_{TA}$ among the early portion and the median contrast in the response to stimulation in B, and associated linear regression

To show this, we sought to isolate this subset of early responders to B. In a given network and stimulus assignment, we found, for each neuron, the minimum response time (i.e., time to first spike) across B trials in \dev{}, \msc{} and \std{} sequences, and ranked neurons by this response time. The ranking in the sample network is shown in Figure \ref{fig6}a. We found that the response developed in an approximately radial pattern, spreading outward in all directions from the stimulus site. We did notice that the early response appeared to gravitate towards the stimulus site for A, but confirmed that this was the case in all sequences separately, indicating a structural rather than dynamic cause (data not shown). We then notionally split the network into an early portion, containing the first 100 neurons to respond to B, and a late portion, containing the remaining 900 neurons. In Figure \ref{fig6}b, we related the non-target response contrast (\dev{} - \msc{}) to the contrast in $V_{TA}$ at the start of B trials. Across excitatory neurons, a strong linear relationship (Pearson's $\rho$ = 0.993, n = 800) echoed the related finding in Figure \ref{fig5}c, where we correlated the same response contrast to $\Delta V_{TA}$ before A. As the color coding and inset histograms show, however, there was a clear separation between the early and late portions of the network, with many early neurons (red) increasing their response in the \dev{} sequence, while the response of the later portion of the network decreased. Correspondingly, the TA levels of the early portion also tended to increase, in clear contrast to the later portion.

@@

As noted above, we hypothesized that, in the full model, STD is directly responsible only for a reduction of non-target \dev{} trials, relative to \msc{}, whereas differences between target \dev{} and \msc{} trials are not a result of STD directly. To show this, we calculated $D$ for A and B trials, and contrasted it across \dev{} and \msc{} sequences as with $V_{TA}$ before.
In Figure \ref{fig8}a, we see that target trials (top) experienced depression across large parts of the sample network in both sequences. The \msc{} sequence, however, clearly evoked greater depression, as borne out in the contrast (\dev{} - \msc{}), which was negative throughout. Conversely, depression in B trials (bottom) was spatially more clustered around the stimulation site in both sequences. In the \msc{} sequence, we can attribute the difference between A and B primarily to different response patterns: The context, and thus $x$, was approximately the same, but responses to B tended not to spread out as much in this network. However, in the \dev{} sequence, the B response was limited almost exclusively to a very local patch around the stimulation site, with depression therefore similarly spatially limited. As a result, the contrast (\dev{} - \msc{}) showed slightly reduced $D$ in the periphery, but strikingly elevated $D$ near the stimulation site, indicating greatly increased synaptic depression in \dev{}.
